Co-evolution of lexical meaning \& pragmatic use

### 2018-03-17 paper/cogsci-submission2.tex
minor edits intro and RMD section

\begin{abstract}
  According to standard linguistic theory, the meaning of an utterance is the product of
  conventional semantic meaning and general pragmatic rules on language use. To investigate how
  cultural evolution of language plays out under this picture of the semantics-pragmatics
  interface, we present a game-theoretic model of the competition between types of language
  users, each endowed with certain lexical representations and a particular pragmatic
  disposition to act on them. Our model traces two evolutionary forces and their interaction:
  (i) pressure towards communicative efficiency and (ii) transmission perturbations during the
  acquisition of linguistic knowledge. We illustrate the model based on a case study on scalar
  implicatures, which suggests that the relationship between underspecified semantics and
  pragmatic inference is one of co-evolution. The evolution of underspecified semantics
  requires pragmatic reasoning, while pragmatic reasoning may not evolve if lexical meanings do
  not require pragmatic enrichment.
\end{abstract}

@@

In contrast, we will here look at language users with a richer linguistic make-up. The objects
of replication and selection in our model are pairs consisting of a set of lexical meanings and
a manner of pragmatic behavior, which together define an agent's linguistic choices. Types of
pragmatic behavior are represented using probabilistic models of pragmatic language use
\citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}. Replication
and selection are described by the \emph{replicator mutator dynamic}, a general and established
model of evolutionary change in large and homogeneous populations
\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,nowak:2006}. This
approach allows us to study the interaction between (i) pressure towards communicative
efficiency and (ii) infidelity in the transmission of linguistic knowledge, caused by factors
such as inductive learning biases and sparse learning data
\citep{KirbyHurford2002:The-Emergence-o,SmithKirby2003:Iterated-Learni,KirbyGriffith2014:Iterated-Learni}. Considering
the effects that the transmission of linguistic knowledge through (iterated) learning have is
important because neither semantic meaning nor pragmatic usage patterns are directly
observable. Instead, language learners have to infer these unobservables from the observable
behavior in which they result. We formalize this process as a form of Bayesian inference. Our
approach thereby contains a well-understood model of iterated Bayesian learning
\citep{griffiths+kalish:2005,griffiths+kalish:2007}, but combines it with functional selection,
here formalized as the most versatile dynamic from evolutionary game theory, the replicator
dynamic \citep{TaylorJonker1978:Evolutionary-St}. Section~\ref{sec:model} introduces this
model.

Section~\ref{sec:si-case-study} applies this model to a case study on scalar implicatures. It
shows that inductive learning biases of Bayesian learners that favor simpler lexical meanings
may play a role in preventing the lexicalization of scalar inferences and thereby lead to the
emergence of Gricean-like pragmatic reasoning types, but only if there is selective pressure
for communicative efficiency and learnability. The results of this study are critically
assessed in the light of the assumptions that feed our model in Section~\ref{sec:discussion}.

@@

The emergence and change of linguistic structure is influenced by many intertwined
factors. These range from biological and socio-ecological to cultural ones \citep{benz+etal:2005b,steels:2011,tamariz+kirby:2016}. Social and ecological pressures determine communicative needs, while
biology determines the architecture that enables and constrains the means by which they can be
fulfilled. In the following, our focus lies on cultural aspects, wherein processes of
linguistic change are viewed as shaped by language use and its transmission, i.e., as a result
of a process of cultural evolution
\citep{Pagel2009:Human-Language-,ThompsonKirby2016:Culture-Shapes-}.

The idea that language is an adaptation to serve a communicative function is fundamental to
many synchronic and diachronic analyses, at least since \citeposs{zipf:1949} explanation of
word frequency rankings as a result of competing hearer and speaker preferences \citep[e.g.,
in][]{martinet:1962, horn:1984,jaeger+vRooij:2007,jaeger:2007,
  piantadosi:2014,kirby+etal:2015}. If processes of selection, such as conditional imitation or
reinforcement, favor more communicatively efficient types of behavior, languages are driven
towards semantic expressivity \citep[e.g.,][]{nowak+krakauer:1999,Skyrms2010:Signals}. But
pressure towards communicative efficiency is not the only force that shapes
language. Learnability is another, as natural languages need to be learnable to survive their
faithful transmission across generations. Furthermore, the effects of even small learning
biases can have striking effects on an evolving language in a process of iterated learning
\citep{KirbyHurford2002:The-Emergence-o,SmithKirby2003:Iterated-Learni,kirby+etal:2014}.

@@

\subsection{The replicator mutator dynamic}
An elegant formal approach to capture the interaction between selection for communicative
efficiency and learnability is the \emph{replicator mutator dynamic}
\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,nowak:2006}. The
RMD is a population dynamic which describes the average change in the composition of a
community of agents of different types. The dynamic assumes a virtually infinite population, so
that it is able to abstract away from what individual agents do. It describes the most probable
trajectory of change in the population vector $\vec{x}$, which codifies relative frequencies of
types, so that $x_i$ is the relative frequency of type $i$ and $\sum\nolimits_i x_i = 1$. In
its simplest discrete-time formulation, the RMD defines the frequency $x'_i$ of each type $i$
in the population at the next time step as a function of: (i) the frequency $x_i$ of each type
$i$ before the update step, (ii) the fitness $f_i$ of each type $i$ before the update, and
(iii) the probability $Q_{ji}$ that an agent who observes the behavior of an agent with type
$j$ ends up acquiring type $i$ (with $\sum_k Q_{jk} = 1$ for all $j$):
\begin{align}
  \label{eq:RMD_discrete}
  x'_i = \sum_j \frac{x_jf_j Q_{ji}}{\Phi}\,,
\end{align}
where $\Phi = \sum_k x_k f_k$ is the average fitness in the population.

@@

The RMD is an abstract description of population-level changes. It abstracts away from specific
assumptions about how agents interact and what causes selection and mutation. To understand
this better, the following looks at each component in isolation: first replication, then
mutation.

According to the replicator dynamic the relative frequency $x_i$ of type $i$ increases
proportional to its average fitness in the population. This dynamic is popular and versatile
because it can be derived from many processes of biological and cultural transmission and
selection \citep[for overview and several derivations
see][]{Sandholm2010:Population-Game}. Under a biological interpretation, the model describes
population-level changes when the number of offspring under single-parent reproduction is
proportional to the parent's fitness. Under a cultural interpretation, examples of concrete
agent-level processes that lead to population-level changes as described by the replicator
dynamic include conditional imitation
\citep[e.g.,][]{Helbing1996:A-Stochastic-Be,Schlag1998:Why-Imitate-and} or reinforcement
learning \citep[e.g.,][]{BorgersSarin997:Learning-Throug,Beggs2005:On-the-Converge}. In
conditional imitation, for instance, each agent carries a fixed type $i$ which it may change to
a type $j$ when it happens to observe another agent using type $j$. Imitation occurs with a
probability that is a monotonic function of $f_j$, i.e., of how well $j$ fares against the
population: the better the behavior $j$, the more likely it is to be imitated.\footnote{A
  crucial assumption in all of the derivations of the replicator dynamic as a biological or
  cultural process of evolution is that agents themselves are rather limited in their strategic
  capacities. They are, for example, not able to innovate a strategy, i.e., to just choose a
  type, that is optimal given the current population state. They are also not able to
  strategically adapt their behavior to any particular agent they are interacting with on some
  given occasion. Such minimalism about the strategic capacities of agents is a conservative
  methodological choice. It allows us to study ---as the formal thought experiment that any
  such model is--- how fitness-based selection changes a population in the absence of optimal
  agent-level reasoning and it abstracts away from specific assumptions about how fitness-based
  selection might be realized concretely.}

If all types have equal fitness at all times, the replicator mutator dynamic reduces to a
process of iterated mutation. Mutation can be interpreted as biological or cultural as
well. Under a biological interpretation where types are genetically encoded behavioral traits,
$Q_{ji}$ is the probability by which offspring of type $j$ genetically mutates into type
$i$. Under a cultural interpretation where types are transmitted by observation or learning,
such as in conditional imitation, $Q_{ji}$ is the probability that an agent trying to learn,
adopt or imitate the behavior of an agent with type $j$ acquires type $i$. As detailed below,
in this way, the process in (\ref{eq:RMD_discrete}), and its equivalent in
(\ref{eq:RMD_discrete_recast}), can contain a model of iterated learning
\citep{griffiths+kalish:2005,griffiths+kalish:2007}.

In the following application to language evolution, we prefer a cultural interpretation of the
mutator component as the perturbation induced by how easy different linguistic traits are
learnable from observation of language use. Nothing of current relevance hinges on whether the
replicator component is interpreted as a biological process or a cultural process. We believe
that this flexibility of interpretation is an asset. It abstracts away from specific
assumptions and focuses on a general and versatile description of fitness-based
selection. Still, to have something concrete in mind when interpreting the following
application, Appendix~\ref{sec:deriv-repl-mutat} derives the RMD from a specific scheme of
agent-level conditional imitation.

>>>>

\begin{abstract}
  According to standard linguistic theory, the meaning of an utterance is the product of
  conventional semantic meaning and general pragmatic rules on language use. We investigate how
  such a division of labor between semantics and pragmatics could evolve under general
  processes of selection and learning. We present a game-theoretic model of the competition
  between types of language users, each endowed with certain lexical representations and a
  particular pragmatic disposition to act on them. Our model traces two evolutionary forces and
  their interaction: (i) pressure towards communicative efficiency and (ii) transmission
  perturbations during the acquisition of linguistic knowledge. We illustrate the model based
  on a case study on scalar implicatures, which suggests that the relationship between
  underspecified semantics and pragmatic inference is one of co-evolution.
\end{abstract}

@@

In contrast, we will here represent explicitly language users' representations of lexical
meanings and their own particular manner of manner of production and interpretation based on
these lexical representation. Different types of pragmatic behavior are represented using
probabilistic models of pragmatic language use
\citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}. Replication
and selection are described by the \emph{replicator mutator dynamic}, a general and established
model of evolutionary change in large and homogeneous populations
\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,nowak:2006}. This
approach allows us to study the interaction between (i) pressure towards communicative
efficiency and (ii) infidelity in the transmission of linguistic knowledge, caused by factors
such as inductive learning biases and sparse learning data
\citep{KirbyHurford2002:The-Emergence-o,SmithKirby2003:Iterated-Learni,KirbyGriffith2014:Iterated-Learni}.

It is important to consider the effects of transmission of linguistic knowledge through
(iterated) learning because neither semantic meaning nor pragmatic usage patterns are directly
observable. Instead, language learners have to infer these unobservables from the observable
behavior in which they result. We formalize this process as a form of Bayesian inference. Our
approach thereby contains a well-understood model of iterated Bayesian learning
\citep{griffiths+kalish:2005,griffiths+kalish:2007}, but combines it with functional selection,
here formalized as the most versatile dynamic from evolutionary game theory, the replicator
dynamic \citep{TaylorJonker1978:Evolutionary-St}. Section~\ref{sec:model} introduces this
model.

Section~\ref{sec:si-case-study} applies this model to a case study on scalar implicatures. It
shows that inductive learning biases of Bayesian learners that favor simpler lexical meanings
may play a role in preventing the lexicalization of scalar inferences and thereby lead to the
emergence of Gricean-like pragmatic reasoning types, but only if there is selective pressure
for communicative efficiency and learnability. The results of this study are critically
assessed in the light of the simplifying assumptions inherent in our abstract model in Section~\ref{sec:discussion}.

@@

The emergence and change of linguistic structure is influenced by many factors in complex ways
\citep{benz+etal:2005b,steels:2011,tamariz+kirby:2016}. Social and ecological pressures
determine communicative needs, while biology determines the architecture that enables and
constrains the means by which they can be fulfilled. In the following, our focus lies on
cultural aspects, wherein processes of linguistic change are viewed as shaped by language use
and its transmission, i.e., as a result of a process of cultural evolution
\citep{Pagel2009:Human-Language-,ThompsonKirby2016:Culture-Shapes-}.

The idea that language is an adaptation to serve a communicative function is fundamental to
many synchronic and diachronic analyses, at least since \citeposs{zipf:1949} explanation of
word frequency rankings as a result of competing hearer and speaker preferences \citep[e.g.,
in][]{martinet:1962, horn:1984,jaeger+vRooij:2007,jaeger:2007,
  piantadosi:2014,kirby+etal:2015}. If processes of selection, such as conditional imitation or
reinforcement, favors communicative efficiency, languages are driven towards semantic
expressivity \citep[e.g.,][]{nowak+krakauer:1999,Skyrms2010:Signals}. But pressure towards
communicative efficiency is not the only force that shapes language. Learnability is another.
Natural languages need to be learnable to survive their faithful transmission across
generations. Furthermore, the effects of even small learning biases can have striking effects
on an evolving language in a process of iterated learning
\citep{KirbyHurford2002:The-Emergence-o,SmithKirby2003:Iterated-Learni,kirby+etal:2014}.

@@

\subsection{The replicator mutator dynamic}
An elegant formal approach to capture the interaction between selection for communicative
efficiency and learnability is the \emph{replicator mutator dynamic}
\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,nowak:2006}. The
RMD is a population dynamic which describes the average change in the composition of a
community of agents of different types. The dynamic assumes a virtually infinite population, so
that it is able to abstract away from what individual agents do. It describes the most probable
trajectory of change in the population vector $\vec{x}$, in which each component $x_i$ gives
the relative frequency of type $i$ and $\sum\nolimits_i x_i = 1$. In its simplest discrete-time
formulation, the RMD defines the frequency $x'_i$ of each type $i$ in the population at the
next time step as a function of: (i) the frequency $x_i$ of each type $i$ before the update
step, (ii) the fitness $f_i$ of each type $i$ before the update, and (iii) the probability
$Q_{ji}$ that an agent who observes the behavior of an agent with type $j$ ends up acquiring
type $i$ (with $\sum_k Q_{jk} = 1$ for all $j$):
\begin{align}
  \label{eq:RMD_discrete}
  x'_i = \sum_j \frac{x_jf_j Q_{ji}}{\Phi}\,,
\end{align}
where $\Phi = \sum_k x_k f_k$ is the average fitness in the population.

@@

The RMD is an abstract description of population-level changes. It abstracts away, to the
extent possible, from specific assumptions about how agents interact and what exactly causes
selection and mutation. To understand this better, the following looks at each component in
isolation: first replication, then mutation.

According to the replicator dynamic the relative frequency $x_i$ of type $i$ increases
proportional to its average fitness in the population. This dynamic is popular and versatile
because it can be derived from many processes of biological and cultural transmission and
selection \citep[for overview and several derivations
see][]{Sandholm2010:Population-Game}. Under a biological interpretation, the model describes
population-level changes when the number of offspring under single-parent reproduction is
proportional to the parent's fitness. Under a cultural interpretation, examples of concrete
agent-level processes that lead to population-level changes as described by the replicator
dynamic include conditional imitation
\citep[e.g.,][]{Helbing1996:A-Stochastic-Be,Schlag1998:Why-Imitate-and} or reinforcement
learning \citep[e.g.,][]{BorgersSarin997:Learning-Throug,Beggs2005:On-the-Converge}. In
conditional imitation, for instance, each agent carries a fixed type $i$ which it may change to
a type $j$ when it happens to observe another agent using type $j$. Imitation occurs with a
probability that is a monotonic function of $f_j$, i.e., of how well $j$ fares against the
population: the better the behavior $j$, the more likely it is to be imitated.\footnote{A
  crucial assumption in all of the derivations of the replicator dynamic as a biological or
  cultural process of evolution is that agents themselves are rather limited in their strategic
  capacities. They are, for example, not able to innovate a strategy, i.e., to just choose a
  type, that is optimal given the current population state. They are also not able to
  strategically adapt their behavior to any particular agent they are interacting with on some
  given occasion. Such minimalism about the strategic capacities of agents is a conservative
  methodological choice. It is one thing to be able to demonstrate that phenomenon $X$ is what
  a rational agent would do. This is useful but may not always be the most satisfactory answer
  to the question how $X$ happened to evolve. It is another thing to show that even if no agent
  was individually capable of seeing $X$ as the rational thing to do, $X$ would still evolve if
  every agent at least sometimes, locally and myopically tried to optimize its behavior, e.g.,
  by imitating what seems to be working well for others \citep{Skyrms2010:Signals}.}

If all types have equal fitness at all times, the replicator mutator dynamic reduces to a
process of iterated mutation. Mutation can be interpreted as biological or cultural as
well. Under a biological interpretation where types are genetically encoded behavioral traits,
$Q_{ji}$ is the probability by which offspring of type $j$ genetically mutates into type
$i$. Under a cultural interpretation where types are transmitted by observation or learning,
such as in conditional imitation, $Q_{ji}$ is the probability that an agent trying to learn,
adopt or imitate the behavior of an agent with type $j$ acquires type $i$.

In the following application to language evolution, we prefer a cultural interpretation of the
mutator component as the perturbation induced by how easy different linguistic traits are
learnable from observation of language use. In this way, the RMD contains a chain of iterated
(Bayesian) learning as a special case
\citep{griffiths+kalish:2005,griffiths+kalish:2007}. Nothing of current relevance hinges on
whether the replicator component is interpreted as a biological process or a cultural
process. We believe that this flexibility of interpretation is an asset. It abstracts away from
specific assumptions and focuses on a general and versatile description of fitness-based
selection. Still, to have something concrete in mind when interpreting the following
application, Appendix~\ref{sec:deriv-repl-mutat} derives the RMD from a specific scheme of
agent-level conditional imitation.
